### 2025-07-24 paper/presenations/slides.qmd

---
title: counterfactual training
subtitle: Update Meeting Dec 2024
author: 
  - name: Patrick Altmeyer
    url: https://www.patalt.org/
  - name: Arie van Deursen
  - name: Cynthia C. S. Liem
institute: Delft University of Technology
date: today
format: 
  beamer:
    theme: Boadilla
  tudelft-revealjs:
    theme: custom.scss
    self-contained: true
    smaller: false
    scrollable: true
    preview-links: auto
    slide-number: true
    transition: slide
    background-transition: fade
    fig-align: center
    html-math-method: mathjax
    include-in-header:
      - text: |
          <script>
          MathJax = {
            options: {
              menuOptions: {
                settings: {
                  assistiveMml: false
                }
              }
            }
          };
          </script>
crossref: 
  prp-title: RQ
  prp-prefix: RQ
classoption: "notheorems"
---

## Status

- *Code base*: In place and streamlined for reproducibility and configuration. 
- *Experiments*: Lots of work done and results largely supportive of idea.
    - Ran into problems on DelftBlue, which has set me back about 2 weeks. 
- *Paper*: Still bare-bones. 
- *ICML*: Potentially still possible to submit something, but this will be rushed and not "finished".

## Problems on Cluster

- Trying to distribute: 
    1. Models/experiments across processes.
    2. For each model/experiment distribute the counterfactual search across processes.
- Out-of-memory issues, data races, ...
- Multi-processing for models & multi-threading for counterfactual search: low CPU efficiency on DelftBlue (jobs get cancelled).

## The Big Oversight

- **Core problem**: `CounterfactualExplanation` objects are *huge*. They store `X` (the input), `y` (the output), etc. in their own memory.
- Can't easily fit `CounterfactualExplanation` objects on cluster memory and pass them around processes. 
- **Good news**: Relatively straightforward to add `FlattenedCE` with gazzilion times lower memory footprint (done!).
- Nested parallelization issue remains and I will not spend more time trying to make it work.

@@

- All counterfactuals at test time generated using *ECCo*.
- Penalty on energy differential increases from *l.* to *r.*

@@

::::{.columns}::::
:::{.column width='50%'}
![Counterfactual explanations for model trained with CT (*ECCo*).](www/moons_ce_ecco.png){#fig-ce-ecco}
:::
:::{.column width='50%'}
![Counterfactual explanations for conventionally trained model.](www/moons_ce_vanilla.png){#fig-ce-vanilla}
:::
::::

@@

## MNIST (Vanilla vs *ECCo*)

::::{.columns}::::
:::{.column width='50%'}
![Faithful counterfactuals for conventionally trained MLP.](www/poc_model.png){#fig-mnist-model}
:::
:::{.column width='50%'}
![Faithful counterfactuals for same MLP with CT (*ECCo*).](www/poc_model_ct_ecco.png){#fig-mnist-model-ct-ecco}
:::
::::

@@

# Planning Ahead

## Planned Contributions

- Focus on enhancing trustworthiness and applicability of small opaque models. 
    - Add more datasets and neuro-tree models (relatively straightforward).
- Do not pitch as state-of-the-art approach to generative modelling similar to JEMs, but rather as an extension of JEMs to XAI and AR. 
- Limit use of image datasets to illustrate arguments:
    - Example: what happens if digit values can only be increased/decreased?

## Timeline

- Code base is already streamlined very well to allow for grid searches with easy configuration.
- Lots of open tasks and questions (see [issue](https://github.com/JuliaTrustworthyAI/CounterfactualTraining.jl/issues/8))
    1. Add MMD to measure plausibility.
    2. Evaluate adversarial robustness and generative capacity. Compare to AT, JEM.
    3. ... 
- Still aiming for ICML to commit to a timeline, but it seems unlikely to be in the shape I would like.

>>>>

---
title: counterfactual training
subtitle: Update Meeting Dec 2024
institute: Delft University of Technology
date: today
format: 
  beamer:
    theme: Boadilla
  tudelft-revealjs:
    theme: custom.scss
    self-contained: true
    smaller: false
    scrollable: true
    preview-links: auto
    slide-number: true
    transition: slide
    background-transition: fade
    fig-align: center
    html-math-method: mathjax
    include-in-header:
      - text: |
          <script>
          MathJax = {
            options: {
              menuOptions: {
                settings: {
                  assistiveMml: false
                }
              }
            }
          };
          </script>
crossref: 
  prp-title: RQ
  prp-prefix: RQ
classoption: "notheorems"
---

@@

- All counterfactuals at test time generated using *ECCCo*.
- Penalty on energy differential increases from *l.* to *r.*

@@

::::{.columns}::::
:::{.column width='50%'}
![Counterfactual explanations for model trained with CT (*ECCCo*).](www/moons_ce_ecco.png){#fig-ce-ecco}
:::
:::{.column width='50%'}
![Counterfactual explanations for conventionally trained model.](www/moons_ce_vanilla.png){#fig-ce-vanilla}
:::
::::

@@

## MNIST (Vanilla vs *ECCCo*)

::::{.columns}::::
:::{.column width='50%'}
![Faithful counterfactuals for conventionally trained MLP.](www/poc_model.png){#fig-mnist-model}
:::
:::{.column width='50%'}
![Faithful counterfactuals for same MLP with CT (*ECCCo*).](www/poc_model_ct_ecco.png){#fig-mnist-model-ct-ecco}
:::
::::

@@

### 2025-07-24 paper/quarto_aaai/paper.tex

\title{Counterfactual Training: Teaching Models Plausible and Actionable
Explanations}
\author{Patrick Altmeyer \and Aleksander Buszydlik \and Arie van
Deursen \and Cynthia C. S. Liem}
\date{}
\begin{document}
\maketitle
\begin{abstract}
We propose a novel training regime termed counterfactual training that
leverages counterfactual explanations to increase the explanatory
capacity of models. Counterfactual explanations have emerged as a
popular post-hoc explanation method for opaque machine learning models:
they inform how factual inputs would need to change in order for a model
to produce some desired output. To be useful in real-word
decision-making systems, counterfactuals should be plausible with
respect to the underlying data and actionable with respect to the
feature mutability constraints. Much existing research has therefore
focused on developing post-hoc methods to generate counterfactuals that
meet these desiderata. In this work, we instead hold models directly
accountable for the desired end goal: counterfactual training employs
counterfactuals during the training phase to minimize the divergence
between learned representations and plausible, actionable explanations.
We demonstrate empirically and theoretically that our proposed method
facilitates training models that deliver inherently desirable
counterfactual explanations and exhibit greatly improved adversarial
robustness.
\end{abstract}
\section{Introduction}\label{introduction}

@@

To assess outcomes with respect to actionability for non-linear models,
we look at the average costs of valid counterfactuals in terms of their
distances from factual starting points. While this an imperfect proxy of
sensitivity, we hypothesize that CT can reduce these costs by teaching
models to seek plausibility with respect to mutable features, much like
we observe in Figure~\ref{fig-poc} in panel (d) compared to (c). We
supplement this analysis with qualitative findings for integrated
gradients \citep{sundararajan2017ig}. Finally, for predictive
performance, we use standard metrics, such as robust accuracy estimated
on adversarially perturbed data using FGSM
\citep{goodfellow2014explaining}.

>>>>

\title{Counterfactual Training: Teaching Models Plausible and Actionable
Explanations}
\author{}
\date{}
\begin{document}
\maketitle
\begin{abstract}
We propose a novel training regime termed counterfactual training that
leverages counterfactual explanations to increase the explanatory
capacity of models. Counterfactual explanations have emerged as a
popular post-hoc explanation method for opaque machine learning models:
they inform how factual inputs would need to change in order for a model
to produce some desired output. To be useful in real-word
decision-making systems, counterfactuals should be plausible with
respect to the underlying data and actionable with respect to the
feature mutability constraints. Much existing research has therefore
focused on developing post-hoc methods to generate counterfactuals that
meet these desiderata. In this work, we instead hold models directly
accountable for the desired end goal: counterfactual training employs
counterfactuals during the training phase to minimize the divergence
between learned representations and plausible, actionable explanations.
We demonstrate empirically and theoretically that our proposed method
facilitates training models that deliver inherently desirable
counterfactual explanations and exhibit greatly improved adversarial
robustness.
\end{abstract}
\section{Introduction}\label{introduction}

@@

To assess outcomes with respect to actionability for non-linear models,
we look at the costs of (just) valid counterfactuals in terms of their
distances from factual starting points with \(\tau=0.5\). While this an
imperfect proxy of sensitivity, we hypothesize that CT can reduce these
costs by teaching models to seek plausibility with respect to mutable
features, much like we observe in Figure~\ref{fig-poc} in panel (d)
compared to (c). We supplement this analysis with estimates for
integrated gradients \citep{sundararajan2017ig}. Finally, for predictive
performance, we use standard metrics, such as robust accuracy estimated
on adversarially perturbed data using FGSM
\citep{goodfellow2014explaining}.

### 2025-07-24 paper/sections/conclusion.qmd

# Conclusions {#sec-conclusions}

State-of-the-art machine learning models are prone to learning complex representations that cannot be interpreted by humans. Existing explainability solutions cannot guarantee that explanations agree with these learned representation. As a step towards addressing this challenge, we introduce counterfactual training, a novel training regime that incentivizes highly-explainable models. Through extensive experiments, we demonstrate that CT satisfies this objective while promoting robustness and preserving the predictive performance of models. Explanations generated from CT-based models are both more plausible (compliant with the underlying data-generating process) and more actionable (compliant with user-specified mutability constraints), and thus meaningful to their recipients. We also show that CT can be used to fine-tune conventionally-trained models and achieve similar gains. Lastly, our work highlights the value of simultaneously improving models and their explanations.

>>>>

# Conclusion {#sec-conclusion}

State-of-the-art machine learning models are prone to learning complex representations that cannot be interpreted by humans. Existing explainability solutions cannot guarantee that explanations agree with these learned representation. As a step towards addressing this challenge, we introduce counterfactual training, a novel training regime that integrates recent advances in contrastive learning, adversarial robustness, and counterfactual explanations to incentivize highly-explainable models. Through extensive experiments, we demonstrate that CT satisfies this goal while preserving the predictive performance and promoting robustness of models. Explanations generated from CT-based models are both more plausible (compliant with the underlying data-generating process) and more actionable (compliant with user-specified mutability constraints), and thus meaningful to their recipients. In turn, our work highlights the value of simultaneously improving models and their explanations.

### 2025-07-24 paper/sections/intro.qmd

Counterfactual explanations (CE) have become a key part of the broader explainable AI (XAI) toolkit [@molnar2022interpretable] that can be applied to make sense of this complexity. Originally proposed by @wachter2017counterfactual, CEs prescribe minimal changes for factual inputs that, if implemented, would prompt some fitted model to produce an alternative, more desirable output (@fig-poc). This is useful and necessary to not only understand how opaque models make their predictions, but also to provide algorithmic recourse (AR) to individuals subjected to them: a retail bank, for example, could use CE to provide meaningful feedback to unsuccessful loan applicants that were rejected based on an opaque automated decision-making (ADM) system.

For such feedback to be meaningful, counterfactual explanations need to fulfill certain desiderata [@verma2020counterfactual; @karimi2020survey]---they should be faithful to the model [@altmeyer2024faithful], plausible [@joshi2019realistic] and actionable [@ustun2019actionable]. Plausibility is typically understood as counterfactuals being *in-domain*: unsuccessful loan applicants that implement the provided recourse should end up with credit profiles that are genuinely similar to that of individuals who have successfully repaid their loans in the past. Actionable explanations comply with practical constraints: a young, unsuccessful loan applicant cannot increase their age in an instant.

Existing state-of-the-art (SOTA) approaches in the field have largely focused on designing model-agnostic CE methods that identify subsets of counterfactuals, which comply with specific desiderata. This is problematic, because the narrow focus on any specific desideratum can adversely affect others: it is possible, for example, to generate plausible counterfactuals for models that are also highly vulnerable to implausible, possibly adversarial counterfactuals [@altmeyer2024faithful]. In this work, we therefore embrace the paradigm that models---as opposed to explanation methods---should be held accountable for explanations that are plausible and actionable. While previous work has shown that at least plausibility can be indirectly achieved through existing techniques aimed at models' generative capacity, generalization and robustness [@altmeyer2024faithful; @augustin2020adversarial; @schut2021generating], we directly incorporate both plausibility and actionability in the training objective of models to improve their overall explanatory capacity.

@@

The remainder of this manuscript is structured as follows. @sec-lit presents related work, focusing on the links to contrastive and robust learning. Then follow our two principal contributions. In @sec-method, we introduce our methodological framework and show theoretically that it can be employed to enforce global actionability constraints. In our experiments (@sec-experiments), we find that thanks to counterfactual training, (1) the implausibility of CEs decreases by up to 90%; (2) the cost of reaching valid counterfactuals with protected features decreases by 19% on average; and (3) models' adversarial robustness improves across the board. Finally, we discuss open challenges in @sec-discussion and conclude in @sec-conclusion.

![Counterfactual explanations (stars) for linear classifiers trained under different regimes on synthetic data: (a) conventional training, all mutable; (b) CT, all mutable; (c) conventional, *age* immutable; (d) CT, *age* immutable. The linear decision boundary is shown in green along with training data colored according to ground-truth labels: $y^-=1$ (blue) and $y^+=2$ (orange). Stars indicate counterfactuals in the target class. For illustration, synthetic features are annotated as *debt* and *age*.](/paper/figures/poc.svg){#fig-poc fig-env="figure*"}

>>>>

Counterfactual explanations (CE) have become a key part of the broader explainable AI (XAI) toolkit [@molnar2022interpretable] that can be applied to make sense of this complexity. Originally proposed by @wachter2017counterfactual, CEs prescribe minimal changes for factual inputs that, if implemented, would prompt some fitted model to produce an alternative, more desirable output. This is useful and necessary to not only understand how opaque models make their predictions, but also to provide algorithmic recourse (AR) to individuals subjected to them: a retail bank, for example, could use CE to provide meaningful feedback to unsuccessful loan applicants that were rejected based on an opaque automated decision-making (ADM) system (@fig-poc).

For such feedback to be meaningful, counterfactual explanations need to fulfill certain desiderata [@verma2020counterfactual; @karimi2020survey]---they should be faithful to the model [@altmeyer2024faithful], plausible [@joshi2019realistic] and actionable [@ustun2019actionable]. Plausibility is typically understood as counterfactuals being *in-domain*: unsuccessful loan applicants that implement the provided recourse should end up with credit profiles that are genuinely similar to that of individuals who have successfully repaid their loans in the past. Actionable explanations comply with practical constraints: a young, unsuccessful loan applicant cannot increase their age in an instance.

Existing state-of-the-art (SOTA) approaches in the field have largely focused on designing model-agnostic CE methods that identify subsets of counterfactuals, which comply with specific desiderata. This is problematic, because the narrow focus on any specific desideratum can adversely affect others: it is possible, for example, to generate plausible counterfactuals for models that are also highly vulnerable to implausible, possibly adversarial counterfactuals [@altmeyer2024faithful]. In this work, we therefore embrace the paradigm that models (as opposed to explanation methods) should be held accountable for explanations that are plausible and actionable. While previous work has shown that at least plausibility can be indirectly achieved through existing techniques aimed at models' generative capacity, generalization and robustness [@altmeyer2024faithful; @augustin2020adversarial; @schut2021generating], we directly incorporate both plausibility and actionability in the training objective of models to improve their overall explanatory capacity.

@@

The remainder of this manuscript is structured as follows. @sec-lit presents related work, focusing on the links to contrastive and robust learning. Then follow our two principal contributions. In @sec-method, we introduce our methodological framework and show theoretically that it can be employed to respect global actionability constraints. In our experiments (@sec-experiments), we find that thanks to counterfactual training, (1) the implausibility of CEs decreases by up to 90%; (2) the cost of reaching valid counterfactuals with protected features decreases by 19% on average; and (3) models' adversarial robustness improves across the board. Finally, we discuss open challenges in @sec-discussion and conclude in @sec-conclusion.

![Counterfactual explanations (stars) for linear classifiers trained under different regimes on synthetic data: (a) conventional training, all mutable; (b) CT, all mutable; (c) conventional, *age* immutable; (d) CT, *age* immutable. The linear decision boundary is shown in green along with training data colored according to ground-truth labels: $y^-=\text{"loan withheld"}$ (blue) and $y^+=\text{"loan provided"}$ (orange). Class and feature annotations (*debt* and *age*) are for illustrative purposes.](/paper/figures/poc.svg){#fig-poc fig-env="figure*"}

### 2025-07-24 paper/sections/other/notation.qmd

@@

- $y^+$: The target class and also the index of the target class.
- $y^-$: The non-target class and also the index of non-the target class.
- $\mathbf{x}$: a single training sample.
- $\mathbf{x}^\prime$: a counterfactual. 
- $\mathbf{x}^+$: a training sample in the target class (ground-truth).
- $\mathbf{y}^+$: The one-hot encoded output vector for the target class. 
- $\theta$: Model parameters (unspecified).
- $\Theta$: Matrix of parameters. 
- $\mathbf{M}(\cdot)$: linear predictions (logits) of the classifier.

## Other Technical Details

@@

>>>>

## Variables and Parameters

@@

- $\mathcal{Y}$: The output domain.
- $y^+$: The target class and also the index of the target class.
- $y^-$: The non-target class and also the index of non-the target class.
- $\mathcal{X}$: The input domain.
- $\mathbf{x}$: a single training sample.
- $\mathbf{x}^\prime$: a counterfactual.
- $t=1,...,T$: Step indicator for counterfactual search iterations.
- $\mathbf{x}^\prime_{t \leq T}$: a nascent counterfactual, defined as a counterfactual that has not yet converged.
- $\mathbf{x}^\prime_{T}$: a mature counterfactual, defined as a counterfactual that has either passed a threshold probability or exhausted all $T$ steps.
- $\mathbf{x}^+$: a training sample in the target class (ground-truth).
- $\mathbf{y}^+$: The one-hot encoded output vector for the target class.
- $\theta$: Model parameters (unspecified).
- $\Theta$: Matrix of parameters. 
- $\mathbf{M}(\cdot)$: linear predictions (logits) of the classifier.

## Formulas

### Maximum Mean Discrepancy

@@

## Other Conventions

In some place of this appendix, we use the terms *full/Full* (i.e. the full CT objective) and *vanilla/Vanilla* (i.e. vanilla training objective) to refer to models trained with counterfactual training (*CT*) and the baseline (*BL*), respectively.
